### 2024-11-06 Report_Template_updated.Rmd
Initial commit

---
title: "Gillespie algorithm for simulating stochastic equations of molecular reactions"
author: "First  Last Name"
date: "`r Sys.Date()`" #  
output:
  pdf_document:
      number_sections: true
spacing: double
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Title of Project}
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
- \usepackage{setspace}\doublespacing
---

```{=tex}
\newpage 
\tableofcontents 
\newpage
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction
This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and 
MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

The rest of this project is as follows. In Section 2, we review some basic concepts. 
In Section 3, we introduce the topic. In Section 4, we have an application and simualtion. 
Section 5 provides conclusion and some closing remarks.

Section 2: original algorithm and basic theory
Section 3: modified algorithm
Section 4: example
Section 5: conclusion (how the modification is better, when to use)

# Review of the Topic

When you click the **Knit** button a document will be generated that includes both content as well 
as the output of any embedded R code chunks within the document. You can embed an R code chunk like 
this:

```{r cars,comment=NA}
summary(cars)
```

You can also embed publication quality plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

## Definitions and Examples
Such as importance of longitudinal data analysis (separate age and cohort effect like in the effect of age on reading ability), different correlation structures, missing data, etc.

## Details of the topic
This is a mathematical formua without formula number.
$$h(x) = \frac{f(x)}{S(x)} = -\frac{d \log[S(x])}{\rm dx}$$

This is how write a fomula and label it. 
\begin{align}
\label{eq2}
S(x) = \exp[-H(x)] = \exp[-\int_0^x h(u) {\rm d}u]
\end{align}

# Application
This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and 
MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

\begin{table}[!htp]
\small
\caption{This is a caption.}
\begin{center}
\begin{tabular}{|c|cc|}
\hline
& {Placebo}   &  {Treatment}       \\ \cline{1-3}  
Method& {Est} (SE)    &{Est} (SE) \\    %  95\% CI \hspace{0.5cm}  95\% CI 
\hline
\scriptsize{Method 1 } & ??? (???)     & ??? (???)      \\   %   (???, ???) \hspace{0.25cm} (???, ???)
\scriptsize{Method 2}  &  ??? (???)     &  ??? (???)     \\  %  (???,  ???)  \hspace{0.25cm}  (???, ???)
\hline
\end{tabular}
\label{tab: 2.3}
\end{center}
\end{table}

# Conclusion

# References
"Lu, Ting, et al. Cellular growth and division in the Gillespie algorithm. Systems biology 1.1 (2004):
121-128."

### 2024-11-06 Report_Template.Rmd
initial template

--- 2024-11-06 Report_Template_updated.Rmd
@@
 ---
-title: "Gillespie algorithm for simulating stochastic equations of molecular reactions"
+title: "Title of the Project"
 author: "First  Last Name"
-date: "`r Sys.Date()`" #  
+date: "`r Sys.Date()`" # or DEcember 1, 2023
 output:
   pdf_document:
-      number_sections: true
-spacing: double
-header-includes:
-- \usepackage{fancyhdr}
-- \pagestyle{fancy}
-- \fancyhead[L]{Title of Project}
-- \fancyfoot[CO,CE]{}
-- \fancyfoot[LE,RO]{\thepage}
-- \usepackage{setspace}\doublespacing
+    toc: yes
+  html_document:
+    number_sections: yes
+    toc: yes
 ---
-
-```{=tex}
-\newpage 
-\tableofcontents 
-\newpage
-```
 
@@
 Section 5 provides conclusion and some closing remarks.
-
-Section 2: original algorithm and basic theory
-Section 3: modified algorithm
-Section 4: example
-Section 5: conclusion (how the modification is better, when to use)
 
@@
 # References
-"Lu, Ting, et al. Cellular growth and division in the Gillespie algorithm. Systems biology 1.1 (2004):
-121-128."

### 2024-11-20 paper.Rmd
Added simulation in application

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex 
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{Title of Project}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}\doublespacing
bibliography: references.bib 
editor_options: 
  markdown: 
    wrap: 72
---

\newpage 
\tableofcontents 
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

This is an R Markdown document. Markdown is a simple formatting syntax
for authoring HTML, PDF, and MS Word documents. For more details on
using R Markdown see <http://rmarkdown.rstudio.com>. We analysis the data in R(@citeR).

The rest of this project is as follows. In Section 2, we review some
basic concepts. In Section 3, we introduce the topic. In Section 4, we
have an application and simualtion. Section 5 provides conclusion and
some closing remarks.

# Review of the Topic

When you click the **Knit** button a document will be generated that
includes both content as well as the output of any embedded R code
chunks within the document. You can embed an R code chunk like this:

```{r cars,comment=NA}
summary(cars)
```

You can also embed publication quality plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

## Definitions and Examples

Such as importance of longitudinal data analysis (separate age and
cohort effect like in the effect of age on reading ability), different
correlation structures, missing data, etc.

## Details of the topic

This is a mathematical formua without formula number.
$$h(x) = \frac{f(x)}{S(x)} = -\frac{d \log[S(x])}{\rm dx}$$

This is how write a fomula and label it. \begin{align}
\label{eq2}
S(x) = \exp[-H(x)] = \exp[-\int_0^x h(u) {\rm d}u]
\end{align}

# Application

```{r}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)

# Define reaction rates
c1 <- (1 - (37 + 2) / 1e6) * 100   # Decay of T2
c2 <- (37 / 1e6) * 100             # Formation of CPD
c3 <- (2 / 1e6) * 100              # Formation of 64-PP

# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)

# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T2"],               # Reaction 1
    c2 * state["T1"] * state["T2"], # Reaction 2
    c3 * state["T2"]                # Reaction 3
  )
}

# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

  # Record the initial state
  initial_row <- c(Time = time, state)
  output <- rbind(output, as.data.frame(t(initial_row), stringsAsFactors = FALSE))

  while (time < max_time) {
    # Calculate propensities
    props <- propensities(state)
    total_prop <- sum(props)

    # Stop if no reactions can occur
    if (total_prop == 0) break

    # Time to next reaction
    delta_t <- rexp(1, rate = total_prop)
    time <- time + delta_t

    # Select which reaction occurs
    reaction <- sample(length(props), 1, prob = props)

    # Update state based on reaction
    if (reaction == 1) state["T2"] <- state["T2"] - 1
    if (reaction == 2) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["CPD"] <- state["CPD"] + 1
    }
    if (reaction == 3) {
      state["T2"] <- state["T2"] - 1
      state["64PP"] <- state["64PP"] + 1
    }

    # Record the new state
    new_row <- c(Time = time, state)
    output <- rbind(output, as.data.frame(t(new_row), stringsAsFactors = FALSE))
  }

  # Convert columns to numeric for consistency
  output[] <- lapply(output, as.numeric)
  return(output)
}

# Run the simulation
max_time <- 1000
simulation <- simulate_gillespie(state, max_time)

```

```{r}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")

# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "Decay-Dimerization Reaction Set Simulation",
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()

```
```{r}
# Display the first few rows of the simulation data
knitr::kable(head(simulation), format = "html", caption = "Simulation Results (First Few Rows)")

```

# Conclusion

# References

### 2024-11-21 paper.Rmd
update

@@
 # Define reaction rates
-c1 <- (1 - (37 + 2) / 1e6) * 100   # Decay of T2
+c1 <- (1 - (37 + 2) / 1e6) * 100   # No dimerization occurs
 c2 <- (37 / 1e6) * 100             # Formation of CPD

### 2024-11-24 paper.Rmd
updated application and references

--- 2024-11-20 paper.Rmd
@@
 # Application
+    We can simulate the dimerization of DNA pyrimidines, a molecular reaction which can lead to skin cancer.
+Bipyrimidine photodimerization occurs when UV radiation alters the chemical bonds of two consecutive pyrimidine bases, with two possible outcomes: the formation of the cyclobutene pyrimidine dimer (CPD) if a cycloaddition between the C5-C6 double bond of the pyrimidines occurs, or the formation of pyrimidine (6-4) pyrimidone (64-PP) if  a covalent bond forms between C4 and C6 of the pyrimidines (@MartinezFernandez2022).
+
+  Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP.
+
+  We will simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per 10^6 normal bases) and 64-PP (~2 per 10^6 normal bases) found by exposing skin cells to a UVB dose of 0.2 J/cm^2 (@Mouret2006).
+As the amount of thymine bases in DNA varies from person to person, we will simulate thymine photodimerization of 1000 pairs of thymine.
 
@@
 # Define reaction rates
-c1 <- (1 - (37 + 2) / 1e6) * 100   # Decay of T2
-c2 <- (37 / 1e6) * 100             # Formation of CPD
-c3 <- (2 / 1e6) * 100              # Formation of 64-PP
+c1 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs
+c2 <- (37 / 1e6) * 10             # Formation of CPD
+c3 <- (2 / 1e6) * 10              # Formation of 64-PP
 
@@
 # Run the simulation
-max_time <- 1000
+max_time <- 100
 simulation <- simulate_gillespie(state, max_time)

### 2024-11-24 paper.Rmd
updated introduction and references

@@
 
-This is an R Markdown document. Markdown is a simple formatting syntax
-for authoring HTML, PDF, and MS Word documents. For more details on
-using R Markdown see <http://rmarkdown.rstudio.com>. We analysis the data in R(@citeR).
+The Gillespie Algorithm was presented in 1976 by Daniel T.
+Gillespie (@Gillespie1976).
+This algorithm was developed  to answer the question, “If a fixed volume V contains a spatially uniform mixture of N chemical species which can inter-react through M specified chemical reaction channels, then given the numbers of molecules of each species present at some initial time, what will these molecular population levels be at any later time?” (@Gillespie1977).
+Given the trajectory of this molecular reaction is a stochastic process, this algorithm can simulate various molecular equation systems, such as radioactive decay models or decaying-dimerization reaction sets (@GillespieSSA).
 
-The rest of this project is as follows. In Section 2, we review some
-basic concepts. In Section 3, we introduce the topic. In Section 4, we
-have an application and simualtion. Section 5 provides conclusion and
-some closing remarks.
+In the following report, we will review basic concepts of the Gillespie Algorithm (Section 2), then we will present the theory of the Gillespie Algorithm (Section 3).
+Afterwards, we will demonstrate a stochastic simulation using the Gillespie algorithm of a dimerization reaction set (Section 4), using R(@citeR).
+Finally, we will provide our concluding remarks (Section 5).
 

### 2024-11-24 paper.Rmd
Added second simulation

@@
 
-```{r}
+```{r simulation_1, echo=FALSE}
 # Load required libraries
-if (!require("reshape2")) install.packages("reshape2")
-if (!require("ggplot2")) install.packages("ggplot2")
 library(reshape2)
@@
 
-```{r}
+```{r graph_1, echo=FALSE}
 # Reshape data for plotting
@@
 ```
-```{r}
-# Display the first few rows of the simulation data
-knitr::kable(head(simulation), format = "html", caption = "Simulation Results (First Few Rows)")
+
+## Radioactive Decay Simulation
+
+This simulation models the stochastic decay of a single species (\( R \)), which represents a radioactive substance such as radon, radium, or plutonium.
+The decay process is governed by the following reaction:
+
+\[
+R \xrightarrow{c} 0
+\]
+
+where \( c \) is the decay rate constant that defines the likelihood of decay per unit time.
+
+The Gillespie algorithm is used to simulate this system, capturing the inherent randomness of radioactive decay events.
+Unlike deterministic approaches, the Gillespie algorithm tracks individual decay events over time, providing an accurate representation of the stochastic nature of the process.
+
+The goal of this simulation is to observe the time evolution of the remaining radioactive molecules (\( R \)) and visualize their decay over time.
+```{r simulation_2, echo=FALSE}
+# Define parameters
+c <- 0.01  # Decay rate constant
+state <- c(R = 1000)  # Initial radioactive substance
+max_time <- 100  # Maximum simulation time
+
+# Define propensity function
+propensities <- function(state) {
+  c * state["R"]
+}
+
+# Gillespie simulation function
+simulate_gillespie <- function(state, max_time) {
+  time <- 0
+  output <- data.frame(Time = numeric(0), R = numeric(0))
+  output <- rbind(output, data.frame(Time = time, R = state["R"]))
+
+  while (time < max_time) {
+    # Calculate propensity
+    prop <- propensities(state)
+    if (prop == 0) break  # Stop if no reactions can occur
+
+    # Time to next reaction
+    delta_t <- rexp(1, rate = prop)
+    time <- time + delta_t
+
+    # Update state (decay reaction)
+    state["R"] <- state["R"] - 1
+
+    # Record the state
+    output <- rbind(output, data.frame(Time = time, R = state["R"]))
+  }
+
+  return(output)
+}
+
+# Run the simulation
+simulation <- simulate_gillespie(state, max_time)
 
 ```
+
+```{r graph_2, echo=FALSE}
+library(ggplot2)
+
+# Plot results
+ggplot(simulation, aes(x = Time, y = R)) +
+  geom_line(color = "blue") +
+  labs(
+    title = "Radioactive Decay Simulation",
+    x = "Time",
+    y = "Remaining Radioactive Substance (R)"
+  ) +
+  theme_minimal()
+
+```
+
+### Observations
+
+1. **Exponential Decay**:
+   - The radioactive substance (\( R \)) decays exponentially over time, as expected from the properties of radioactive decay.
+The number of decay events per unit time decreases as the quantity of \( R \) diminishes.
+   - This behavior is reflected in the propensity function \( a(R) = c \cdot R \), where the decay rate is proportional to the remaining amount of \( R \).
+
+2. **Stochastic Fluctuations**:
+   - Unlike a smooth deterministic decay curve, the simulation shows small fluctuations in the timing of decay events.
+These fluctuations arise from the stochastic nature of the Gillespie algorithm, which models each decay event as a probabilistic process.
+
+3. **Time Steps and Decay Events**:
+   - The time intervals between decay events (\( \Delta t \)) grow longer as the number of \( R \) molecules decreases.
+This is because the propensity (\( a(R) \)) becomes smaller, leading to larger expected time intervals between events.
+
+### Biological/Physical Significance
+
+- **Radioactive Half-Life**:
+  - The exponential decay observed in this simulation is consistent with the concept of half-life in radioactive decay.
+While this model does not explicitly compute the half-life, it demonstrates how the decay rate slows down as the substance depletes.
+
+- **Real-World Applications**:
+  - Radioactive decay processes are critical in many fields, including nuclear physics, medicine (e.g., radiotherapy), and environmental science (e.g., radon gas monitoring).
+Understanding stochastic decay helps refine predictions in low-count systems where randomness plays a significant role.
 

### 2024-11-25 paper.Rmd
made change

--- 2024-11-24 paper.Rmd
@@
     number_sections: true
-    citation_package: biblatex 
 spacing: double
@@
   - \usepackage{setspace}\doublespacing
-bibliography: references.bib 
 editor_options: 
@@
 # Application
+
 We can simulate the dimerization of DNA pyrimidines, a molecular reaction which can lead to skin cancer.

### 2024-11-25 paper.Rmd
Uploaded APA References

--- 2024-11-24 paper.Rmd
@@
 
-In the following report, we will review basic concepts of the Gillespie Algorithm (Section 2), then we will present the theory of the Gillespie Algorithm (Section 3).
-Afterwards, we will demonstrate a stochastic simulation using the Gillespie algorithm of a dimerization reaction set (Section 4), using R(@citeR).
-Finally, we will provide our concluding remarks (Section 5).
+In the following report, we will review basic concepts of the Gillespie Algorithm [Topic Review section](#sec-Topic-Review), then we will present the theory of the Gillespie Algorithm [Theory section](#sec-theory).
+Afterwards, we will demonstrate a stochastic simulation using the Gillespie algorithm of a dimerization reaction set [APplication section](#sec-application), using R(@citeR).
+Finally, we will provide our concluding remarks [Conclusion section](#sec-conclusion).
 
-# Review of the Topic
+# Topic Review {#sec-Topic-Review}
 
@@
 
-# Application
+# Theory {#sec-theory}
+# Application {#sec-application}
+## Photodimerization Simulation
 We can simulate the dimerization of DNA pyrimidines, a molecular reaction which can lead to skin cancer.
@@
 
-```{r simulation_1, echo=FALSE}
+```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
 # Load required libraries
+if (!require("reshape2")) install.packages("reshape2")
+if (!require("ggplot2")) install.packages("ggplot2")
 library(reshape2)
 library(ggplot2)
-
 # Define reaction rates
@@
 c3 <- (2 / 1e6) * 10              # Formation of 64-PP
-
 # Initial state
 state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
-
 # Define propensity functions
@@
 }
-
 # Gillespie simulation function
@@
 }
-
 # Run the simulation
@@
 simulation <- simulate_gillespie(state, max_time)
-
 ```
 
-```{r graph_1, echo=FALSE}
+```{r graph_1, echo=FALSE, warning=FALSE, message=FALSE}
 # Reshape data for plotting
 simulation_long <- melt(simulation, id.vars = "Time")
-
 # Plot results
@@
   theme_minimal()
-
 ```
@@
 ## Radioactive Decay Simulation
-
 This simulation models the stochastic decay of a single species (\( R \)), which represents a radioactive substance such as radon, radium, or plutonium.
 The decay process is governed by the following reaction:
-
 \[
@@
 \]
-
 where \( c \) is the decay rate constant that defines the likelihood of decay per unit time.
-
 The Gillespie algorithm is used to simulate this system, capturing the inherent randomness of radioactive decay events.
 Unlike deterministic approaches, the Gillespie algorithm tracks individual decay events over time, providing an accurate representation of the stochastic nature of the process.
-
 The goal of this simulation is to observe the time evolution of the remaining radioactive molecules (\( R \)) and visualize their decay over time.
-```{r simulation_2, echo=FALSE}
+```{r simulation_2, echo=FALSE, warning=FALSE, message=FALSE}
 # Define parameters
@@
 max_time <- 100  # Maximum simulation time
-
 # Define propensity function
@@
 }
-
 # Run the simulation
 simulation <- simulate_gillespie(state, max_time)
-
 ```
-
 ```{r graph_2, echo=FALSE}
 library(ggplot2)
-
 # Plot results
@@
 ### Observations
-
 1. **Exponential Decay**:
@@
    - This behavior is reflected in the propensity function \( a(R) = c \cdot R \), where the decay rate is proportional to the remaining amount of \( R \).
-
 2. **Stochastic Fluctuations**:
@@
 These fluctuations arise from the stochastic nature of the Gillespie algorithm, which models each decay event as a probabilistic process.
-
 3. **Time Steps and Decay Events**:
@@
 ### Biological/Physical Significance
-
 - **Radioactive Half-Life**:
@@
 While this model does not explicitly compute the half-life, it demonstrates how the decay rate slows down as the substance depletes.
-
 - **Real-World Applications**:
@@
 
-# Conclusion
+# Conclusion {#sec-conclusion}
 

### 2024-11-26 test.Rmd
test rmd for updated dimerization simulation; can be adjusted to run faster w/o reaction 3

```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
#SLOW COMPUTATION.. FOR FAST COMPUTE SEE LINES 11, 20, 65, & 79 
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (37 / 1e6) * 10             # Formation of CPD
c2 <- (2 / 1e6) * 10              # Formation of 64-PP
c3 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs// REMOVE THIS LINE FOR FAST COMPUTE

# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T1"] * state["T2"], # Reaction 1
    c2 * state["T1"] * state["T2"],  # Reaction 2
    c3 * state["T1"] * state["T2"]  #Reaction 3   // REMOVE THIS LINE FOR FAST COMPUTE
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

  # Record the initial state
  initial_row <- c(Time = time, state)
  output <- rbind(output, as.data.frame(t(initial_row), stringsAsFactors = FALSE))

  while (time < max_time) {
    # Calculate propensities
    props <- propensities(state)
    total_prop <- sum(props)

    # Stop if no reactions can occur
    if (total_prop == 0) break

    # Time to next reaction
    delta_t <- rexp(1, rate = total_prop)
    time <- time + delta_t

    # Select which reaction occurs
    reaction <- sample(length(props), 1, prob = props)

    # Update state based on reaction
    if (reaction == 1) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["CPD"] <- state["CPD"] + 1
    }
    if (reaction == 2) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["64PP"] <- state["64PP"] + 1
    }
    if (reaction == 3) {   #REMOVE THIS SECTION FOR QUICK COMPUTE
      next
    }

    # Record the new state
    new_row <- c(Time = time, state)
    output <- rbind(output, as.data.frame(t(new_row), stringsAsFactors = FALSE))
  }

  # Convert columns to numeric for consistency
  output[] <- lapply(output, as.numeric)
  return(output)
}
# Run the simulation
max_time <- 10        #SET TO HIGHER TIME IF DOING FAST COMPUTE
simulation <- simulate_gillespie(state, max_time)
```

```{r graph_1, echo=FALSE, warning=FALSE, message=FALSE}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")
# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "Dimerization Reaction Set Simulation",
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()
```
```{r}
# Display the first few rows of the simulation data
knitr::kable(head(simulation), format = "html", caption = "Simulation Results (First Few Rows)")
```

### 2024-11-26 test.Rmd
added rough observations to test.rmd

@@
 ```
+###Observations
+1.**Logarithmic Growth**
+Both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (Martinez-Fernandez and Improta, 2022; Mouret et al., 2006).
+The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
+
+2. **Stochastic Fluctuations**
+Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.
+
+3. **Time Steps and Reaction Events**
+
+###Biological Significance
+-**Photodimerization**
+As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers.
+-**Real-World Application**
+Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer. 
+This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels.

### 2024-11-26 paper.Rmd
Edited second simulation

@@
 output:
-  pdf_document:
     number_sections: true
@@
 ## Radioactive Decay Simulation
+
 This simulation models the stochastic decay of a single species (\( R \)), which represents a radioactive substance such as radon, radium, or plutonium.
 The decay process is governed by the following reaction:
+
 \[
@@
 \]
+
 where \( c \) is the decay rate constant that defines the likelihood of decay per unit time.
+
 The Gillespie algorithm is used to simulate this system, capturing the inherent randomness of radioactive decay events.
@@
 The goal of this simulation is to observe the time evolution of the remaining radioactive molecules (\( R \)) and visualize their decay over time.
+
 ```{r simulation_2, echo=FALSE, warning=FALSE, message=FALSE}
@@
 c <- 0.01  # Decay rate constant
-state <- c(R = 1000)  # Initial radioactive substance
-max_time <- 100  # Maximum simulation time
+state <- c(R = 10000)  # Initial radioactive substance
+max_time <- 1000  # Extended maximum simulation time
+
 # Define propensity function
@@
 }
+
 # Gillespie simulation function
@@
 }
+
 # Run the simulation
 simulation <- simulate_gillespie(state, max_time)
+
 ```
-```{r graph_2, echo=FALSE}
+
+```{r decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}
+
+# Load ggplot2 for visualization
 library(ggplot2)
-# Plot results
+
+# Linear-scale plot
 ggplot(simulation, aes(x = Time, y = R)) +
@@
   labs(
-    title = "Radioactive Decay Simulation",
     x = "Time",
@@
 
+```{r decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
+# Logarithmic-scale plot
+ggplot(simulation, aes(x = Time, y = R)) +
+  geom_line(color = "red") +
+  scale_y_log10() +
+  labs(
+    x = "Time",
+    y = "Remaining Radioactive Substance (log(R))"
+  ) +
+  theme_minimal()
+
+```
+
 ### Observations
-1. **Exponential Decay**:
-   - The radioactive substance (\( R \)) decays exponentially over time, as expected from the properties of radioactive decay.
-The number of decay events per unit time decreases as the quantity of \( R \) diminishes.
-   - This behavior is reflected in the propensity function \( a(R) = c \cdot R \), where the decay rate is proportional to the remaining amount of \( R \).
-2. **Stochastic Fluctuations**:
-   - Unlike a smooth deterministic decay curve, the simulation shows small fluctuations in the timing of decay events.
-These fluctuations arise from the stochastic nature of the Gillespie algorithm, which models each decay event as a probabilistic process.
-3. **Time Steps and Decay Events**:
-   - The time intervals between decay events (\( \Delta t \)) grow longer as the number of \( R \) molecules decreases.
-This is because the propensity (\( a(R) \)) becomes smaller, leading to larger expected time intervals between events.
 
-### Biological/Physical Significance
-- **Radioactive Half-Life**:
-  - The exponential decay observed in this simulation is consistent with the concept of half-life in radioactive decay.
-While this model does not explicitly compute the half-life, it demonstrates how the decay rate slows down as the substance depletes.
-- **Real-World Applications**:
-  - Radioactive decay processes are critical in many fields, including nuclear physics, medicine (e.g., radiotherapy), and environmental science (e.g., radon gas monitoring).
-Understanding stochastic decay helps refine predictions in low-count systems where randomness plays a significant role.
+The simulation results are visualized in two plots.
+Figure \@ref(fig:decay-linear) shows the decay on a linear scale, while Figure \@ref(fig:decay-log) presents the same data on a logarithmic scale to highlight the exponential nature of the decay process.
+
+#### Exponential Decay
+The radioactive substance (\( R \)) decays exponentially over time, as expected from the properties of radioactive decay.
+In Figure \@ref(fig-decay-linear), this behavior is reflected in the steep decline in \( R \) during the early stages of the simulation, followed by a more gradual decline as the remaining quantity of \( R \) decreases.
+This is consistent with the exponential decay law:
+
+\[
+R(t) = R_0 e^{-ct}
+\]
+
+where \( R_0 \) is the initial amount, \( c \) is the decay constant, and \( t \) is time.
+The propensity function \( a(R) = c \cdot R \) ensures that the decay rate is proportional to the remaining amount of \( R \).
+
+In Figure \@ref(fig-decay-log), the decay trajectory appears as a straight line, which confirms the exponential nature of the process.
+The slope of this line corresponds to the decay constant \( c \), validating the theoretical prediction.
+
+#### Stochastic Fluctuations
+Unlike deterministic simulations, the stochastic approach introduces minor fluctuations in the timing of individual decay events.
+These fluctuations are more noticeable when the number of molecules is small, as random variations in the timing of events have a larger relative impact.
+
+#### Time Steps and Decay Events
+As the number of molecules decreases, the propensity \( a(R) \) becomes smaller, leading to larger intervals (\( \Delta t \)) between successive decay events.
+This dynamic is evident in both figures, where the rate of decline slows significantly in the later stages of the simulation.
+
+### Biological and Physical Significance
+
+#### Radioactive Half-Life
+The exponential decay observed in this simulation aligns with the concept of half-life, the time required for half of the substance to decay. The half-life (\( t_{1/2} \)) can be computed as:
+
+\[
+t_{1/2} = \frac{\ln(2)}{c}
+\]
+
+For \( c = 0.01 \), the half-life is approximately \( t_{1/2} \approx 69.3 \) time units.
+This simulation demonstrates how the decay rate diminishes over time as the remaining radioactive substance depletes.
+
+#### Real-World Applications
+Radioactive decay is a fundamental process in nuclear physics, medicine, and environmental science.
+For example, it is critical in radiotherapy for cancer treatment, where the controlled decay of radioactive isotopes delivers targeted doses of radiation.
+It also has applications in monitoring environmental contaminants such as radon gas, where stochastic decay processes affect detection and measurement accuracy.
+Understanding the inherent randomness of decay is essential for accurate predictions, especially in low-count systems where stochastic effects dominate.
 

### 2024-11-27 paper.Rmd
9:12pm adding theory

--- 2024-11-25 paper.Rmd
@@
 Finally, we will provide our concluding remarks [Conclusion section](#sec-conclusion).
+
+## background
+In this report we studied the original version of the Gillespie algorithm.The Gillespie algorithm generates a statistically correct trajectory to simulate chemical or biochemical systems of reactions efficiently .
+It allows us to model dynamics that can’t be solved by the master equation analytically, especially for complicated examples.
+
+## introduction
+Biochemical reactions in cells are inherently stochastic :reactions can occur at any moment, and at any given time, the state of the system determines the probability of a particular chemical reaction occurring next.
+This is due to the random thermal motion of molecules and the small numbers of reactant molecules involved.
+When molecule counts are low (often dozens or hundreds rather than millions), random fluctuations in molecule collisions and interactions become significant, making the timing and order of reactions probabilistic rather than deterministic.
+
+## theory
+## Master Equation
+
+Generally we use the distribution function $P(S,t)$ to describe a many particle system at time t, where $P(S,t)$ is the probability for the system to be in the state $S$ at time $t$.
+
+The equation governing the evolution of $P(S,t)$ is called the master equation. During the time interval $[t,t+dt)$, $P(S,t)$ is changed due to two types of process:
+
+1. The particles in the state $S$ leave due to some reaction $S\rightarrow S'$.
+
+2. The particles in other state $S'$ enter the state $S$ state due to some reaction $S'\rightarrow S$.
+
+Combining the two factors, we obtain the master equation with the form $^{[1]}$ $$\frac{\partial}{\partial t}P(S,t)=\sum_{S'}P(S',t)R(S',S)-\sum_{S'}P(S,t)R(S,S')$$, where $R(S,S')$ represents the reaction rate from the state $S$ to the state $S'$.
+
+The idea of master equation is inherently stochastic.
+Firstly $P(S,t)$ itself is a probabilistic description of the system.
+Next, in most of the cases, the reaction/process is stochastic.
+Therefore the master equation describes the ensemble average of all possible systems, here `ensemble` is the collection of all possible states with the same initial condition and macroscopic parameters.
+Following some certain initial distribution, we are not guaranteed to reach the same distribution at time $t$.
+
+In the article `Cellular growth and division in the Gillespie algorithm` $^{[2]}$, the authors discussed the system of well-stirred mixture of $N$ chemical species.
+Since the system is well-stirred, the distribution of particles is always uniform in space.
+The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system.
+Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$.
+Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$.
+Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.
+
+For example, in a second order reaction $A+B\rightarrow AB$, we have $$\frac{d}{dt}X_{AB}=V\frac{d}{dt}[AB]=V*c'[A][B]=\frac{c'}{V}*X_AX_B=c*h(Y)$$, where $c'$ is the true reaction constant measured in chemistry, $V$ is the volume of the system and $[i]=\frac{X_i}{V}$ is the number concentration of species $i$.
+
+With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$, here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.
+
+## Gillespie Algorithm
+
+The master equation is a coupled linear ordinary equation for $|S_Y|$ dependent variables $P(Y,t|Y_0,t_0)$, where $|S_Y|$ is the size of the state space.
+For a system with $n$ particles, $|S_Y|$ is of order $O(n^N)$.
+Therefore it is almost impossible to solve the equation analytically, especially when the number of possible elementary reactions $M$ is large.
+
+A practical alternative to solve the master equation is the stochastic simulation approach devised by Gillespie $^{[3]}$.
+Using the idea that $P(Y,t|Y_0,t_0)$ describes the ensemble average of all possible micro-states given initial state $Y_0$ at time $t_0$, the Gillespie algorithm generates an ensemble of sample trajectories of the system with statistics which asymptotically converges to the solution of the corresponding Master Equation.
+
+In the Gillespie algorithm, we update the state of the system by determining
+
+(i) the time $\tau$ to the next reaction
+
+(ii) which reaction $R_{\mu}$ will occur next.
+
+Assuming that each reaction $R_{\mu}$ is independent with rate $a_{\mu}=h_{\mu}c_{\mu}$, the first occurrence $\tau_{\mu}$ of the reaction $R_{\mu}$ follows the exponential distribution with rate/propensity $a_{\mu}$.
+The first occurrence time of some reaction is then $\tau=\min_{\mu}\tau_{\mu}$, it follows  the exponential distribution with an overall rate/propensity $A=\sum_{\mu}a_{\mu}$.
+Given that a reaction occurs, the probability that it is the $\mu'$-th reaction is $P(\mu=\mu')=\frac{a_{\mu'}}{A}$.
+Therefore the the original Gillespie recipe follows the following steps:
+
+1) Input values $c_{\mu}$, $\mu=1,...,M$ and initial state $Y_0=(x_1,...,x_N)$ at $t_0$.
+
+2) Compute the current propensities $a_{\mu}=h_{\mu}(Y)c_{\mu}$, $\mu=1,...,M$ and $A=\sum_{\mu}a_{\mu}$
+
+3) Generate uniform random numbers $u_1,u_2\in[0,1)$
+
+4) Compute the time interval $\tau$ until the next reaction according to distribution $\exp(A)$, i.e., $\tau=-\ln u_1/A$
+
+5) Find the channel of the next reaction $\mu$, i.e., take $\mu$ to be the integer for which $\sum_{\nu=1}^{\mu-1}a_{\nu}<u_2*A\le \sum_{\nu=1}^{\mu}a_{\nu}$
+
+6) Update time $t\rightarrow t+\tau$, and adjust $Y$ in accordance with the particular reaction $R_{\mu}$, i.e., update $Y\rightarrow Y+a_{\mu}$, and proceed to step 2 until $t$ reaches presetting terminal time $T_f$ and obtain a final state $Y_f$ at $T_f$.
+
+After generating sufficiently many $N_{Traj}$ such trajectories, we can estimate $P(Y,T_f|Y_0,t_0)=\frac{\# Y_{f,Traj}=Y }{N_{Traj}}$.
+
+The Gillespie algorithm is straightforward and easy to implement.
+However, it is only an approximate rather than an exact stochastic algorithm.
+A very large number of trajectories is needed to reach a reasonable accuracy.
+The complexity of the algorithm is very high and the accuracy is not guaranteed due to randomness.
 
@@
 # References
+## Reference
+
+[1]. N. G. V. Kampen, Stochastic Processes in Physics and Chemistry. North-Holland, Amsterdam, 1992.
+
+[2]. T. Lu, D. Volfson, L. Tsimring, J. Hasty, Cellular growth and division in the Gillespie algorithm, Syst Biol (Stevenage). 2004 Jun;1(1):121-8
+
+[3]. D. T. Gillespie, “Exact stochastic simulation of coupled chemical reactions,” J. Phys. Chem., vol. 81, no. 25, pp. 2340–2361,
+1977.

### 2024-11-27 paper.Rmd
added abstract, conclution, and made change to the simulation

@@
 date: "`r Sys.Date()`"
+abstract: "The Gillespie Algorithm provides a robust framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes.
+Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics.
+This paper explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes.
+Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes.
+The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
 output:
+  pdf_document:
     number_sections: true
@@
   - \pagestyle{fancy}
-  - \fancyhead[L]{Title of Project}
+  - \fancyhead[L]{The Gillespie Algorithm}
   - \fancyfoot[CO,CE]{}
   - \fancyfoot[LE,RO]{\thepage}
-  - \usepackage{setspace}\doublespacing
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage{xcolor} 
+  - \usepackage{hyperref}
+  - \hypersetup{colorlinks = true, linkcolor = blue}
+
 bibliography: references.bib 
@@
 
-In the following report, we will review basic concepts of the Gillespie Algorithm [Topic Review section](#sec-Topic-Review), then we will present the theory of the Gillespie Algorithm [Theory section](#sec-theory).
-Afterwards, we will demonstrate a stochastic simulation using the Gillespie algorithm of a dimerization reaction set [APplication section](#sec-application), using R(@citeR).
-Finally, we will provide our concluding remarks [Conclusion section](#sec-conclusion).
+Biochemical reactions in living cells are inherently stochastic.
+Reactions occur at random intervals, and the state of the system at any given time determines the probability of a particular chemical reaction occurring next.
+This stochasticity arises from the random thermal motion of molecules and the small numbers of reactant molecules involved.
+When molecule counts are low (often in the range of dozens or hundreds), random fluctuations in molecule collisions and interactions become significant, making reaction timing and outcomes probabilistic rather than deterministic.
+Consequently, deterministic approaches fail to accurately capture the dynamics of such systems, necessitating stochastic simulation methods like the Gillespie Algorithm.
 
-## background
-In this report we studied the original version of the Gillespie algorithm.The Gillespie algorithm generates a statistically correct trajectory to simulate chemical or biochemical systems of reactions efficiently .
-It allows us to model dynamics that can’t be solved by the master equation analytically, especially for complicated examples.
+In the following paper, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Topic Review section](#sec-Topic-Review).
+Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Theory section](#sec-Theory).
+We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Application section](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006).
+These simulations will be performed using the R programming language (@citeR).
+Finally, our [Conclusion section](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.
 
-## introduction
-Biochemical reactions in cells are inherently stochastic :reactions can occur at any moment, and at any given time, the state of the system determines the probability of a particular chemical reaction occurring next.
-This is due to the random thermal motion of molecules and the small numbers of reactant molecules involved.
-When molecule counts are low (often dozens or hundreds rather than millions), random fluctuations in molecule collisions and interactions become significant, making the timing and order of reactions probabilistic rather than deterministic.
+# Topic Review {#sec-Topic-Review}
 
-## theory
+## Background
+
+In this paper, we focus on the original version of the Gillespie Algorithm.
+This algorithm generates statistically correct trajectories to simulate chemical or biochemical reaction systems with high efficiency.
+By leveraging stochastic simulation, the Gillespie Algorithm enables researchers to model dynamics that cannot be solved analytically using the master equation, particularly for complex reaction networks.
+Its application is especially valuable in cases where deterministic methods fail to capture the inherent randomness of biochemical processes, such as systems with low molecular counts or highly interconnected reaction pathways.
+
+# Theory {#sec-Theory}
+
 ## Master Equation
@@
 
-# Topic Review {#sec-Topic-Review}
-
-When you click the **Knit** button a document will be generated that
-includes both content as well as the output of any embedded R code
-chunks within the document. You can embed an R code chunk like this:
-
-```{r cars,comment=NA}
-summary(cars)
-```
-
-You can also embed publication quality plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-## Definitions and Examples
-
-Such as importance of longitudinal data analysis (separate age and
-cohort effect like in the effect of age on reading ability), different
-correlation structures, missing data, etc.
-
-## Details of the topic
-
-This is a mathematical formua without formula number.
-$$h(x) = \frac{f(x)}{S(x)} = -\frac{d \log[S(x])}{\rm dx}$$
-
-This is how write a fomula and label it. \begin{align}
-\label{eq2}
-S(x) = \exp[-H(x)] = \exp[-\int_0^x h(u) {\rm d}u]
-\end{align}
-
-# Theory {#sec-theory}
 # Application {#sec-application}
@@
 
+### Observations
+
+The simulation results are presented in two figures, each offering unique insights into the behavior of the decay process.
+Figure \@ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time.
+Figure \@ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.
+
+#### General Explanation of Exponential Decay
+The radioactive decay of \( R \) follows the well-known exponential decay law:
+
+\[
+R(t) = R_0 e^{-ct}
+\]
+
+where \( R_0 \) is the initial quantity of the substance, \( c \) is the decay constant, and \( t \) is time.
+As \( R \) decreases, the propensity function \( a(R) = c \cdot R \) ensures that the decay rate becomes slower over time.
+This results in a high number of decay events at the beginning of the simulation, which gradually diminishes as fewer molecules remain.
+
+While the decay trajectory conforms to deterministic exponential decay theory, the Gillespie algorithm introduces stochastic variations.
+These arise because the time between successive reactions is governed by random exponential waiting times, and the occurrence of each reaction is probabilistic.
+As the number of \( R \) molecules declines, stochastic fluctuations become more apparent due to the reduced scale of the system.
+
 ```{r simulation_2, echo=FALSE, warning=FALSE, message=FALSE}
@@
 
+#### Analysis of Figure \@ref(fig-decay-linear)
+Figure \@ref(fig-decay-linear) illustrates the exponential decay of \( R \) over time.
+The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum.
+As \( R \) decreases, the curve flattens, reflecting the slower decay rate.
+This behavior exemplifies the exponential decay dynamics, where the number of remaining molecules reduces rapidly at first and progressively lessens as the substance depletes.
+
+The gradual flattening of the curve also underscores the diminishing probability of decay events as time progresses.
+This effect is a direct result of the propensity function \( a(R) = c \cdot R \), which decreases linearly with \( R \).
+The linear-scale plot effectively illustrates the overall trend of radioactive decay over time, making it particularly suitable for understanding the macroscopic behavior of the process.
+
 ```{r decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}
@@
 
+\newpage
+#### Analysis of Figure \@ref(fig-decay-log)
+Figure \@ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective.
+On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line.
+The slope of the line corresponds to the negative decay constant, \(-c\), providing a quantitative validation of the theoretical model.
+
+The logarithmic representation also highlights deviations from perfect linearity, which become more noticeable as the molecule count approaches zero.
+These deviations are the result of stochastic fluctuations inherent in the Gillespie algorithm, particularly in the later stages when \( R \) is low, and individual decay events have a greater impact on the overall trend.
+This logarithmic plot is especially useful for analyzing the detailed kinetics of the decay process and for confirming the exponential nature of the decay.
+
 ```{r decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
@@
 
-### Observations
-
-The simulation results are visualized in two plots.
-Figure \@ref(fig:decay-linear) shows the decay on a linear scale, while Figure \@ref(fig:decay-log) presents the same data on a logarithmic scale to highlight the exponential nature of the decay process.
-
-#### Exponential Decay
-The radioactive substance (\( R \)) decays exponentially over time, as expected from the properties of radioactive decay.
-In Figure \@ref(fig-decay-linear), this behavior is reflected in the steep decline in \( R \) during the early stages of the simulation, followed by a more gradual decline as the remaining quantity of \( R \) decreases.
-This is consistent with the exponential decay law:
-
-\[
-R(t) = R_0 e^{-ct}
-\]
-
-where \( R_0 \) is the initial amount, \( c \) is the decay constant, and \( t \) is time.
-The propensity function \( a(R) = c \cdot R \) ensures that the decay rate is proportional to the remaining amount of \( R \).
-
-In Figure \@ref(fig-decay-log), the decay trajectory appears as a straight line, which confirms the exponential nature of the process.
-The slope of this line corresponds to the decay constant \( c \), validating the theoretical prediction.
-
-#### Stochastic Fluctuations
-Unlike deterministic simulations, the stochastic approach introduces minor fluctuations in the timing of individual decay events.
-These fluctuations are more noticeable when the number of molecules is small, as random variations in the timing of events have a larger relative impact.
-
-#### Time Steps and Decay Events
-As the number of molecules decreases, the propensity \( a(R) \) becomes smaller, leading to larger intervals (\( \Delta t \)) between successive decay events.
-This dynamic is evident in both figures, where the rate of decline slows significantly in the later stages of the simulation.
-
 ### Biological and Physical Significance
@@
 #### Radioactive Half-Life
-The exponential decay observed in this simulation aligns with the concept of half-life, the time required for half of the substance to decay. The half-life (\( t_{1/2} \)) can be computed as:
+The exponential decay observed in the simulation aligns with the concept of half-life, a critical measure in the study of radioactive processes.
+The half-life (\( t_{1/2} \)) is defined as the time required for half of the radioactive substance to decay.
+It can be calculated as:
 
@@
 
-For \( c = 0.01 \), the half-life is approximately \( t_{1/2} \approx 69.3 \) time units.
-This simulation demonstrates how the decay rate diminishes over time as the remaining radioactive substance depletes.
+For the decay constant \( c = 0.01 \) used in this simulation, the half-life is approximately \( t_{1/2} \approx 69.3 \) time units.
+This simulation demonstrates how the decay rate slows down over time as the radioactive substance depletes, with the time between successive decay events increasing in proportion to the decreasing molecule count.
 
-#### Real-World Applications
-Radioactive decay is a fundamental process in nuclear physics, medicine, and environmental science.
-For example, it is critical in radiotherapy for cancer treatment, where the controlled decay of radioactive isotopes delivers targeted doses of radiation.
-It also has applications in monitoring environmental contaminants such as radon gas, where stochastic decay processes affect detection and measurement accuracy.
-Understanding the inherent randomness of decay is essential for accurate predictions, especially in low-count systems where stochastic effects dominate.
+#### Implications and Real-World Applications
+Radioactive decay is a fundamental phenomenon in nuclear physics and has wide-ranging applications across multiple scientific disciplines.
+For instance, in nuclear medicine, the controlled decay of radioactive isotopes is utilized in radiotherapy to deliver targeted doses of radiation for cancer treatment.
+Similarly, in environmental science, the decay of isotopes such as radon is monitored to assess contamination levels and potential health risks.
+
+The stochastic approach modeled here provides critical insights into systems with low molecule counts, where randomness plays a significant role.
+For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery.
+Likewise, in environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.
+
+In the broader context of molecular reaction networks, the Gillespie algorithm's ability to capture both macroscopic trends and stochastic variations makes it an invaluable tool for studying systems with inherent randomness.
+By accurately simulating individual reaction events, it bridges the gap between theoretical models and real-world observations, offering a deeper understanding of complex kinetic processes.
 
 # Conclusion {#sec-conclusion}
+The Gillespie Algorithm represents a significant advancement in the numerical simulation of stochastic molecular reaction networks.
+Its ability to generate statistically valid trajectories of reaction systems provides critical insights into processes governed by inherent randomness, especially in systems with low molecule counts.
+This study has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.
+
+The simulation of thymine photodimerization highlighted the algorithm’s capacity to model complex reaction networks, offering a detailed view of the formation of UV-induced DNA lesions such as CPDs and 64-PPs.
+This has direct implications for understanding the molecular origins of UV-induced mutagenesis and skin cancer.
+Similarly, the radioactive decay simulation demonstrated the algorithm’s precision in modeling exponential decay processes and capturing stochastic fluctuations, a crucial feature for understanding radioactive isotopes in fields such as nuclear medicine and environmental science.
+
+While the Gillespie Algorithm excels at modeling stochastic dynamics, it is not without limitations.
+The computational cost of simulating systems with large molecule counts or numerous reaction channels can become prohibitive, and its reliance on random number generation introduces variability that requires a sufficiently large number of trajectories for robust statistical analysis.
+Nevertheless, its utility in studying molecular systems with inherent randomness cannot be overstated.
+
+The implications of this work extend beyond the specific applications discussed.
+By providing a framework for understanding and simulating stochastic reaction systems, the Gillespie Algorithm continues to play a pivotal role in advancing our knowledge of molecular biology, chemical kinetics, and physical chemistry.
+Future developments in computational efficiency and hybrid methods may further expand its applicability, enabling the simulation of even more complex systems with greater precision.
 
 # References
-## Reference
 

### 2024-11-27 paper.Rmd
updated Dimerization Simulation in paper.rmd

@@
 Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP. 
+\[
+T1 + T2 \xrightarrow{c1} CPD
+\]
+\[
+T1 + T2 \xrightarrow{c2} 64PP
+\]
+\[
+T1 + T2 \xrightarrow{c3} T1+T2
+\]
 
@@
 ```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
+#SLOW COMPUTATION.. FOR FAST COMPUTE SEE LINES 11, 20, 65, & 79 
 # Load required libraries
@@
 # Define reaction rates
-c1 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs
-c2 <- (37 / 1e6) * 10             # Formation of CPD
-c3 <- (2 / 1e6) * 10              # Formation of 64-PP
+c1 <- (37 / 1e6) * 10             # Formation of CPD
+c2 <- (2 / 1e6) * 10              # Formation of 64-PP
+c3 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs// REMOVE THIS LINE FOR FAST COMPUTE
+
 # Initial state
@@
   c(
-    c1 * state["T2"],               # Reaction 1
+    c1 * state["T1"] * state["T2"], # Reaction 1
     c2 * state["T1"] * state["T2"],  # Reaction 2
-    c3 * state["T2"]                # Reaction 3
+    c3 * state["T1"] * state["T2"]  #Reaction 3   // REMOVE THIS LINE FOR FAST COMPUTE
   )
@@
     # Update state based on reaction
-    if (reaction == 1) state["T2"] <- state["T2"] - 1
-    if (reaction == 2) {
+    if (reaction == 1) {
       state["T1"] <- state["T1"] - 1
@@
     }
-    if (reaction == 3) {
+    if (reaction == 2) {
+      state["T1"] <- state["T1"] - 1
       state["T2"] <- state["T2"] - 1
       state["64PP"] <- state["64PP"] + 1
+    }
+    if (reaction == 3) {   #REMOVE THIS SECTION FOR QUICK COMPUTE
+      next
     }
@@
 # Run the simulation
-max_time <- 100
+max_time <- 10        #SET TO HIGHER TIME IF DOING FAST COMPUTE
 simulation <- simulate_gillespie(state, max_time)
@@
   labs(
-    title = "Decay-Dimerization Reaction Set Simulation",
+    title = "Dimerization Reaction Set Simulation",
     x = "Time",
@@
 ```
+###Observations
+Both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (Martinez-Fernandez and Improta, 2022; Mouret et al., 2006).
+The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
+Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.
+
+###Biological Significance
+-**Dimerization**
+As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers.
+Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).
+-**Real-World Application**
+Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time.
+This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels, with applications in biology, chemistry, and bioinformatics.
 

### 2024-11-27 paper.Rmd
update

@@
 ```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
-#SLOW COMPUTATION.. FOR FAST COMPUTE SEE LINES 11, 20, 65, & 79 
+#SLOW COMPUTATION.. FOR FAST COMPUTE MAKE ADJUSTMENTS AS NEEDED
 # Load required libraries

### 2024-11-27 paper.Rmd
update

--- 2024-11-27 paper.Rmd
@@
 ---
-
 \newpage
@@
 
-Combining the two factors, we obtain the master equation with the form $^{[1]}$ $$\frac{\partial}{\partial t}P(S,t)=\sum_{S'}P(S',t)R(S',S)-\sum_{S'}P(S,t)R(S,S')$$, where $R(S,S')$ represents the reaction rate from the state $S$ to the state $S'$.
+Combining the two factors, we obtain the master equation with the form (@vanKampen1992) $$\frac{\partial}{\partial t}P(S,t)=\sum_{S'}P(S',t)R(S',S)-\sum_{S'}P(S,t)R(S,S')$$ where $R(S,S')$ represents the reaction rate from the state $S$ to the state $S'$.
 
@@
 
-In the article `Cellular growth and division in the Gillespie algorithm` $^{[2]}$, the authors discussed the system of well-stirred mixture of $N$ chemical species.
+In the article `Cellular growth and division in the Gillespie algorithm` (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species.
 Since the system is well-stirred, the distribution of particles is always uniform in space.
@@
 
-For example, in a second order reaction $A+B\rightarrow AB$, we have $$\frac{d}{dt}X_{AB}=V\frac{d}{dt}[AB]=V*c'[A][B]=\frac{c'}{V}*X_AX_B=c*h(Y)$$, where $c'$ is the true reaction constant measured in chemistry, $V$ is the volume of the system and $[i]=\frac{X_i}{V}$ is the number concentration of species $i$.
+For example, in a second order reaction $A+B\rightarrow AB$, we have $$\frac{d}{dt}X_{AB}=V\frac{d}{dt}[AB]=V*c'[A][B]=\frac{c'}{V}*X_AX_B=c*h(Y)$$ where $c'$ is the true reaction constant measured in chemistry, $V$ is the volume of the system and $[i]=\frac{X_i}{V}$ is the number concentration of species $i$.
 
-With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$, here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.
+With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$ here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.
 
@@
 
-A practical alternative to solve the master equation is the stochastic simulation approach devised by Gillespie $^{[3]}$.
+A practical alternative to solve the master equation is the stochastic simulation approach devised by Gillespie (@Gillespie1977).
 Using the idea that $P(Y,t|Y_0,t_0)$ describes the ensemble average of all possible micro-states given initial state $Y_0$ at time $t_0$, the Gillespie algorithm generates an ensemble of sample trajectories of the system with statistics which asymptotically converges to the solution of the corresponding Master Equation.
@@
 Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP. 
+\[
+T1 + T2 \xrightarrow{c1} CPD
+\]
+\[
+T1 + T2 \xrightarrow{c2} 64PP
+\]
 
@@
 # Define reaction rates
-c1 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs
-c2 <- (37 / 1e6) * 10             # Formation of CPD
-c3 <- (2 / 1e6) * 10              # Formation of 64-PP
+c1 <- (37 / 1e6) * 10             # Formation of CPD
+c2 <- (2 / 1e6) * 10              # Formation of 64-PP
 # Initial state
@@
   c(
-    c1 * state["T2"],               # Reaction 1
-    c2 * state["T1"] * state["T2"], # Reaction 2
-    c3 * state["T2"]                # Reaction 3
+    c1 * state["T1"] * state["T2"], # Reaction 1
+    c2 * state["T1"] * state["T2"]  # Reaction 2
   )
@@
     }
-    if (reaction == 3) {
+    if (reaction == 2) {
+      state["T1"] <- state["T1"] - 1
       state["T2"] <- state["T2"] - 1
@@
 
-```{r graph_1, echo=FALSE, warning=FALSE, message=FALSE}
+```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Simulation"}
 # Reshape data for plotting
@@
 ```
+
+### Observations
+
+Both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006).
+The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
+Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.
+
+### Biological Significance
+- **Dimerization**
+As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers.
+Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).
+- **Real-World Application**
+Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time.
+This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels, with applications in biology, chemistry, and bioinformatics.
 
@@
 
-```{r decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}
+```{r fig-decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}
 
@@
 
-```{r decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
+```{r fig-decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
 # Logarithmic-scale plot
@@
 # References
-
-[1]. N. G. V. Kampen, Stochastic Processes in Physics and Chemistry. North-Holland, Amsterdam, 1992.
-
-[2]. T. Lu, D. Volfson, L. Tsimring, J. Hasty, Cellular growth and division in the Gillespie algorithm, Syst Biol (Stevenage). 2004 Jun;1(1):121-8
-
-[3]. D. T. Gillespie, “Exact stochastic simulation of coupled chemical reactions,” J. Phys. Chem., vol. 81, no. 25, pp. 2340–2361,
-1977.

### 2024-11-27 paper.Rmd
update applicat8ion

@@
 # Application {#sec-application}
-## Photodimerization Simulation
+
+This section explores the practical application of the Gillespie Algorithm by simulating two molecular systems.
+The first simulation, [Photodimerization Simulation](#sec-Photodimerization), models the photodimerization of thymine, a molecular reaction relevant to DNA damage caused by ultraviolet (UV) radiation.
+The second simulation, [Radioactive Decay Simulation](#sec-Decay), examines the stochastic decay of radioactive isotopes, a fundamental process in nuclear physics.
+Both simulations references the `GillespieSSA` package (@GillespieSSA), a R implementation of the Gillespie Algorithm that facilitates the generation of statistically accurate trajectories for molecular reaction systems.
+By consulting this package, we can simulate complex stochastic processes and explore the dynamics of these reaction systems.
+
+## Photodimerization Simulation {#sec-Photodimerization}
 We can simulate the dimerization of DNA pyrimidines, a molecular reaction which can lead to skin cancer.
@@
 
-## Radioactive Decay Simulation
+## Radioactive Decay Simulation {#sec-Decay}
 

### 2024-11-27 paper.Rmd
updated with prof's edits

@@
 Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics.
-This paper explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes.
+This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes.
 Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes.
@@
 
-In the following paper, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Topic Review section](#sec-Topic-Review).
-Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Theory section](#sec-Theory).
-We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Application section](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006).
+In the following sections, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Section 2](#sec-Topic-Review).
+Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory).
+We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006).
 These simulations will be performed using the R programming language (@citeR).
-Finally, our [Conclusion section](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.
+Finally, our [Section 5](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.
 
@@
 
-In this paper, we focus on the original version of the Gillespie Algorithm.
+In this project, we focus on the original version of the Gillespie Algorithm.
 This algorithm generates statistically correct trajectories to simulate chemical or biochemical reaction systems with high efficiency.
@@
 
-In the article `Cellular growth and division in the Gillespie algorithm` (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species.
+In the article *Cellular growth and division in the Gillespie algorithm* (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species.
 Since the system is well-stirred, the distribution of particles is always uniform in space.
@@
 The simulation results are presented in two figures, each offering unique insights into the behavior of the decay process.
-Figure \@ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time.
-Figure \@ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.
+Figure \ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time.
+Figure \ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.
 
@@
 
-#### Analysis of Figure \@ref(fig-decay-linear)
-Figure \@ref(fig-decay-linear) illustrates the exponential decay of \( R \) over time.
+#### Analysis of Figure \ref(fig-decay-linear)
+Figure \ref{fig:fig-decay-linear} illustrates the exponential decay of \( R \) over time.
 The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum.
@@
 \newpage
-#### Analysis of Figure \@ref(fig-decay-log)
-Figure \@ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective.
+#### Analysis of Figure \ref(fig-decay-log)
+ Figure \ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective.
 On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line.

### 2024-11-27 paper.Rmd
fixed Figure cross-references

@@
 
-```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Simulation"}
+```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Reaction Set Simulation \\label{fig:dimerization-sim}"}
 # Reshape data for plotting
@@
   labs(
-    title = "Dimerization Reaction Set Simulation",
     x = "Time",
@@
 The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
-Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.
+Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction (Figure \ref{fig:dimerization-sim}).
 
@@
 The simulation results are presented in two figures, each offering unique insights into the behavior of the decay process.
-Figure \ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time.
-Figure \ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.
+Figure \ref{fig:decay-linear} displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time.
+Figure \ref{fig:decay-log} presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.
 
@@
 
-#### Analysis of Figure \ref(fig-decay-linear)
-Figure \ref{fig:fig-decay-linear} illustrates the exponential decay of \( R \) over time.
+#### Analysis of Figure \ref{fig:decay-linear}
+Figure \ref{fig:decay-linear} illustrates the exponential decay of \( R \) over time.
 The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum.
@@
 
-```{r fig-decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}
+```{r fig-decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale \\label{fig:decay-linear}"}
 
@@
 \newpage
-#### Analysis of Figure \ref(fig-decay-log)
- Figure \ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective.
+#### Analysis of Figure \ref{fig:decay-log}
+ Figure \ref{fig:decay-log} depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective.
 On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line.
@@
 
-```{r fig-decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
+```{r fig-decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale \\label{fig:decay-log}"}
 # Logarithmic-scale plot

### 2024-11-27 test.Rmd
new test for colouring links

--- 2024-11-26 test.Rmd
@@
-```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
-#SLOW COMPUTATION.. FOR FAST COMPUTE SEE LINES 11, 20, 65, & 79 
-# Load required libraries
-if (!require("reshape2")) install.packages("reshape2")
-if (!require("ggplot2")) install.packages("ggplot2")
-library(reshape2)
-library(ggplot2)
-# Define reaction rates
-c1 <- (37 / 1e6) * 10             # Formation of CPD
-c2 <- (2 / 1e6) * 10              # Formation of 64-PP
-c3 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs// REMOVE THIS LINE FOR FAST COMPUTE
+---
+title: "Test Document"
+output:
+  pdf_document:
+    keep_tex: true
+header-includes:
+  - \usepackage{xcolor}
+  - \usepackage{hyperref}
+  - \hypersetup{colorlinks=true, linkcolor=red, citecolor=green, urlcolor=blue}
+---
 
-# Initial state
-state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
-# Define propensity functions
-propensities <- function(state) {
-  c(
-    c1 * state["T1"] * state["T2"], # Reaction 1
-    c2 * state["T1"] * state["T2"],  # Reaction 2
-    c3 * state["T1"] * state["T2"]  #Reaction 3   // REMOVE THIS LINE FOR FAST COMPUTE
-  )
-}
-# Gillespie simulation function
-simulate_gillespie <- function(state, max_time) {
-  # Initialize time and output data frame
-  time <- 0
-  output <- data.frame(
-    Time = numeric(0),
-    T1 = numeric(0),
-    T2 = numeric(0),
-    CPD = numeric(0),
-    `64PP` = numeric(0)
-  )
+This is a test document. See Figure \ref{fig:test}.
 
-  # Record the initial state
-  initial_row <- c(Time = time, state)
-  output <- rbind(output, as.data.frame(t(initial_row), stringsAsFactors = FALSE))
-
-  while (time < max_time) {
-    # Calculate propensities
-    props <- propensities(state)
-    total_prop <- sum(props)
-
-    # Stop if no reactions can occur
-    if (total_prop == 0) break
-
-    # Time to next reaction
-    delta_t <- rexp(1, rate = total_prop)
-    time <- time + delta_t
-
-    # Select which reaction occurs
-    reaction <- sample(length(props), 1, prob = props)
-
-    # Update state based on reaction
-    if (reaction == 1) {
-      state["T1"] <- state["T1"] - 1
-      state["T2"] <- state["T2"] - 1
-      state["CPD"] <- state["CPD"] + 1
-    }
-    if (reaction == 2) {
-      state["T1"] <- state["T1"] - 1
-      state["T2"] <- state["T2"] - 1
-      state["64PP"] <- state["64PP"] + 1
-    }
-    if (reaction == 3) {   #REMOVE THIS SECTION FOR QUICK COMPUTE
-      next
-    }
-
-    # Record the new state
-    new_row <- c(Time = time, state)
-    output <- rbind(output, as.data.frame(t(new_row), stringsAsFactors = FALSE))
-  }
-
-  # Convert columns to numeric for consistency
-  output[] <- lapply(output, as.numeric)
-  return(output)
-}
-# Run the simulation
-max_time <- 10        #SET TO HIGHER TIME IF DOING FAST COMPUTE
-simulation <- simulate_gillespie(state, max_time)
-```
-
-```{r graph_1, echo=FALSE, warning=FALSE, message=FALSE}
-# Reshape data for plotting
-simulation_long <- melt(simulation, id.vars = "Time")
-# Plot results
-ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
-  geom_line() +
-  labs(
-    title = "Dimerization Reaction Set Simulation",
-    x = "Time",
-    y = "Molecule Count",
-    color = "Species"
-  ) +
-  theme_minimal()
-```
-```{r}
-# Display the first few rows of the simulation data
-knitr::kable(head(simulation), format = "html", caption = "Simulation Results (First Few Rows)")
-```
+```{r fig-test, echo=FALSE, fig.cap="Test Figure \\label{fig:test}"}
+plot(1:10)

### 2024-11-27 paper.Rmd
added Appendix

@@
 # References
+
+\newpage
+
+# Appendix
+
+## R Code for [4.1 Photodimerization Simulation](#sec-Photodimerization)
+
+```{r a1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
+# Load required libraries
+if (!require("reshape2")) install.packages("reshape2")
+if (!require("ggplot2")) install.packages("ggplot2")
+library(reshape2)
+library(ggplot2)
+# Define reaction rates
+c1 <- (37 / 1e6) * 10             # Formation of CPD
+c2 <- (2 / 1e6) * 10              # Formation of 64-PP
+# Initial state
+state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
+# Define propensity functions
+propensities <- function(state) {
+  c(
+    c1 * state["T1"] * state["T2"], # Reaction 1
+    c2 * state["T1"] * state["T2"]  # Reaction 2
+  )
+}
+# Gillespie simulation function
+simulate_gillespie <- function(state, max_time) {
+  # Initialize time and output data frame
+  time <- 0
+  output <- data.frame(
+    Time = numeric(0),
+    T1 = numeric(0),
+    T2 = numeric(0),
+    CPD = numeric(0),
+    `64PP` = numeric(0)
+  )
+
+  # Record the initial state
+  initial_row <- c(Time = time, state)
+  output <- rbind(output, as.data.frame(t(initial_row), stringsAsFactors = FALSE))
+
+  while (time < max_time) {
+    # Calculate propensities
+    props <- propensities(state)
+    total_prop <- sum(props)
+
+    # Stop if no reactions can occur
+    if (total_prop == 0) break
+
+    # Time to next reaction
+    delta_t <- rexp(1, rate = total_prop)
+    time <- time + delta_t
+
+    # Select which reaction occurs
+    reaction <- sample(length(props), 1, prob = props)
+
+    # Update state based on reaction
+    if (reaction == 1) {
+      state["T1"] <- state["T1"] - 1
+      state["T2"] <- state["T2"] - 1
+      state["CPD"] <- state["CPD"] + 1
+    }
+    if (reaction == 2) {
+      state["T1"] <- state["T1"] - 1
+      state["T2"] <- state["T2"] - 1
+      state["64PP"] <- state["64PP"] + 1
+    }
+
+    # Record the new state
+    new_row <- c(Time = time, state)
+    output <- rbind(output, as.data.frame(t(new_row), stringsAsFactors = FALSE))
+  }
+
+  # Convert columns to numeric for consistency
+  output[] <- lapply(output, as.numeric)
+  return(output)
+}
+# Run the simulation
+max_time <- 100
+simulation <- simulate_gillespie(state, max_time)
+
+# Reshape data for plotting
+simulation_long <- melt(simulation, id.vars = "Time")
+# Plot results
+ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
+  geom_line() +
+  labs(
+    x = "Time",
+    y = "Molecule Count",
+    color = "Species"
+  ) +
+  theme_minimal()
+```
+
+## R code for [4.2 Radioactive Decay Simulation](#sec-Decay)
+```{r a2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
+# Define parameters
+c <- 0.01  # Decay rate constant
+state <- c(R = 10000)  # Initial radioactive substance
+max_time <- 1000  # Extended maximum simulation time
+
+# Define propensity function
+propensities <- function(state) {
+  c * state["R"]
+}
+
+# Gillespie simulation function
+simulate_gillespie <- function(state, max_time) {
+  time <- 0
+  output <- data.frame(Time = numeric(0), R = numeric(0))
+  output <- rbind(output, data.frame(Time = time, R = state["R"]))
+
+  while (time < max_time) {
+    # Calculate propensity
+    prop <- propensities(state)
+    if (prop == 0) break  # Stop if no reactions can occur
+
+    # Time to next reaction
+    delta_t <- rexp(1, rate = prop)
+    time <- time + delta_t
+
+    # Update state (decay reaction)
+    state["R"] <- state["R"] - 1
+
+    # Record the state
+    output <- rbind(output, data.frame(Time = time, R = state["R"]))
+  }
+
+  return(output)
+}
+
+# Run the simulation
+simulation <- simulate_gillespie(state, max_time)
+
+# Load ggplot2 for visualization
+library(ggplot2)
+
+# Linear-scale plot
+ggplot(simulation, aes(x = Time, y = R)) +
+  geom_line(color = "blue") +
+  labs(
+    x = "Time",
+    y = "Remaining Radioactive Substance (R)"
+  ) +
+  theme_minimal()
+
+# Logarithmic-scale plot
+ggplot(simulation, aes(x = Time, y = R)) +
+  geom_line(color = "red") +
+  scale_y_log10() +
+  labs(
+    x = "Time",
+    y = "Remaining Radioactive Substance (log(R))"
+  ) +
+  theme_minimal()
+
+```

### 2024-11-27 paper.Rmd
grammar fix

@@
 The second simulation, [Radioactive Decay Simulation](#sec-Decay), examines the stochastic decay of radioactive isotopes, a fundamental process in nuclear physics.
-Both simulations references the `GillespieSSA` package (@GillespieSSA), a R implementation of the Gillespie Algorithm that facilitates the generation of statistically accurate trajectories for molecular reaction systems.
+Both simulations reference the `GillespieSSA` package (@GillespieSSA), a R implementation of the Gillespie Algorithm that facilitates the generation of statistically accurate trajectories for molecular reaction systems.
 By consulting this package, we can simulate complex stochastic processes and explore the dynamics of these reaction systems.

### 2024-11-27 paper.Rmd
fixed formatting of references and appendix
updated for uniform formatting

@@
 
-Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP. 
+Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T1 \)), Thymine 2 (\( T2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reactions channels: \( c1 \) and \( c2 \).
 \[
@@
 
-```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Reaction Set Simulation \\label{fig:dimerization-sim}"}
+```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Reaction Set Simulation \\label{fig:dimerization}"}
 # Reshape data for plotting
@@
 
-Both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006).
+As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006).
 The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
-Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction (Figure \ref{fig:dimerization-sim}).
+Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.
 
-### Biological Significance
-- **Dimerization**
+### Biological and Physical Significance
+
+#### Dimerization
+
 As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers.
 Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).
-- **Real-World Application**
-Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time.
-This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels, with applications in biology, chemistry, and bioinformatics.
+
+#### Implications and Real-World Applications
+Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time, rather than following a general logarithmic curve.This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels.
+Using the Gillespie algorithm to model molecular reactions, such as the dimerization of thymine shown in this simulation, has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology.
+By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes in disease development, therapy, and environmental impact studies.
 
@@
 
+\newpage
 # References
+<div id="refs"></div>
+\printbibliography[heading=none]
+\def\printbibliography{}
 
@@
 
-# Appendix
-
+\appendix
+\section{Appendix}
 ## R Code for [4.1 Photodimerization Simulation](#sec-Photodimerization)

### 2024-11-27 paper.Rmd
updated text in sim 1

@@
 
-As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006).
-The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
-Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.
+As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, which aligns with expectations based on specific UV radiation conditions and previously reported quantum yields (@MartinezFernandez2022; @Mouret2006).
+The number of reactions per unit time levels off as pyrimidine pairs decrease, which is consistent with the propensity functions, as the propensity of 64-PP formation is much significantly than the propensity of CPD formation (@Mouret2006).
+This plateau suggests that, as the number of available reactants decreases, the rate of reaction slows down.
+Due to the stochastic nature of molecular reactions, this simulation demonstrates how the Gillespie algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each individual reaction event.
+These fluctuations are especially important in low-molecule systems, where randomness significantly impacts the overall reaction dynamics, providing a more accurate representation of the molecular behavior compared to deterministic models.
 
@@
 As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers.
-Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).
+Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates dependent on the monomers involved as well as the level of UV radiation (@Mouret2006).
 
 #### Implications and Real-World Applications
-Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time, rather than following a general logarithmic curve.This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels.
-Using the Gillespie algorithm to model molecular reactions, such as the dimerization of thymine shown in this simulation, has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology.
-By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes in disease development, therapy, and environmental impact studies.
+The Gillespie algorithm offers profound implications for understanding molecular reactions, particularly in systems where stochastic effects are prominent, such as the dimerization.
+This process, highlighted in Figure \ref{fig:dimerization}, is widely studied, as it can lead to mutations potentially resulting in skin cancer, when not repaired.
+The Gillespie  algorithm allows us to realistically model these reactions, as it demonstrates the probabilistic nature of DNA damage, capturing fluctuations in molecular interactions that deterministic models cannot, providing a deeper understanding of molecular events, insights into mutation rates, and the impact of environmental factors like UV radiation on genetic integrity.
+The algorithm's ability to model these stochastic processes is crucial in areas such as genetics, where random molecular events can significantly influence disease outcomes.
+This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels.
+Using the Gillespie algorithm to model molecular reactions has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology.
+By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes, such as in disease development, medicine, and environmental impact studies.
 

### 2024-11-27 paper.Rmd
updated citations

@@
 
-where \( R_0 \) is the initial quantity of the substance, \( c \) is the decay constant, and \( t \) is time.
+where \( R_0 \) is the initial quantity of the substance, \( c \) is the decay constant, and \( t \) is time (@Henriksen2002).
 As \( R \) decreases, the propensity function \( a(R) = c \cdot R \) ensures that the decay rate becomes slower over time.
@@
 The exponential decay observed in the simulation aligns with the concept of half-life, a critical measure in the study of radioactive processes.
-The half-life (\( t_{1/2} \)) is defined as the time required for half of the radioactive substance to decay.
+The half-life (\( t_{1/2} \)) is defined as the time required for half of the radioactive substance to decay (@Henriksen2002).
 It can be calculated as:

### 2024-11-28 paper.Rmd
updated some formatting

@@
 
-Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T1 \)), Thymine 2 (\( T2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reactions channels: \( c1 \) and \( c2 \).
+Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T_1 \)), Thymine 2 (\( T_2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reactions channels: \( c_1 \) and \( c_2 \).
 \[
-T1 + T2 \xrightarrow{c1} CPD
+T_1 + T_2 \xrightarrow{c_1} CPD
 \]
 \[
-T1 + T2 \xrightarrow{c2} 64PP
+T_1 + T_2 \xrightarrow{c_2} 64PP
 \]
 
-We will simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per 10^6 normal bases) and 64-PP (~2 per 10^6 normal bases) found by exposing skin cells to a UVB dose of 0.2 J/cm^2 (@Mouret2006).
+We will simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per \(10^6\) normal bases) and 64-PP (~2 per \(10^6\) normal bases) found by exposing skin cells to a UVB dose of 0.2 \(J/cm^2\) (@Mouret2006).
 As the amount of thymine bases in DNA varies from person to person, we will simulate thymine photodimerization of 1000 pairs of thymine.

### 2024-11-28 paper.Rmd
added colours

@@
     citation_package: biblatex
+linkcolor: blue
+urlcolor: blue
 spacing: double
@@
   - \doublespacing
-  - \usepackage{xcolor} 
-  - \usepackage{hyperref}
-  - \hypersetup{colorlinks = true, linkcolor = blue}
 

### 2024-11-28 test.Rmd
added colours

@@
   pdf_document:
+    latex_engine: xelatex
     keep_tex: true
+    number_sections: true
+    citation_package: biblatex
+linkcolor: blue
+spacing: double
 header-includes:
+  - \AtBeginDocument{\hypersetup{colorlinks=true,linkcolor=blue,filecolor=magenta,urlcolor=cyan}}
+  - \usepackage{fancyhdr}
+  - \pagestyle{fancy}
+  - \fancyhead[L]{The Gillespie Algorithm}
+  - \fancyfoot[CO,CE]{}
+  - \fancyfoot[LE,RO]{\thepage}
+  - \usepackage{setspace}
+  - \doublespacing
+  - \PassOptionsToPackage{hyphens}{url}  % Ensure URLs are broken correctly
   - \usepackage{xcolor} 
   - \usepackage{hyperref}
-  - \hypersetup{colorlinks=true, linkcolor=red, citecolor=green, urlcolor=blue}
+  - \hypersetup{colorlinks=true,linkcolor=blue,filecolor=magenta,urlcolor=cyan}
 ---

### 2024-11-28 paper/paper.Rmd
clean files

--- 2024-11-28 paper.Rmd
@@
   - Anna Szydlowski
+thanks: "We would like to thank Dr.
+Omidali A.
+Jazifor for his invaluable guidance throughout this project.
+The source code for simulations can be found on our [GitHub repository](https://github.com/masinan123/Gillespie-algorithm.git)."
 date: "`r Sys.Date()`"
-abstract: "The Gillespie Algorithm provides a robust framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes.
+abstract: "The Gillespie Algorithm provides a framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes.
 Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics.
@@
 #### General Explanation of Exponential Decay
+
 The radioactive decay of \( R \) follows the well-known exponential decay law:
@@
 
-#### Analysis of Figure \ref{fig:decay-linear}
+#### Analysis of Radioactive Decay Simulation in Linear Scale (Figure \ref{fig:decay-linear})
+
 Figure \ref{fig:decay-linear} illustrates the exponential decay of \( R \) over time.
@@
 \newpage
-#### Analysis of Figure \ref{fig:decay-log}
+#### Analysis of Radioactive Decay Simulation in Logarithmic Scale (Figure \ref{fig:decay-log})
+
 Figure \ref{fig:decay-log} depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective.
@@
 #### Radioactive Half-Life
+
 The exponential decay observed in the simulation aligns with the concept of half-life, a critical measure in the study of radioactive processes.
@@
 #### Implications and Real-World Applications
+
 Radioactive decay is a fundamental phenomenon in nuclear physics and has wide-ranging applications across multiple scientific disciplines.
-For instance, in nuclear medicine, the controlled decay of radioactive isotopes is utilized in radiotherapy to deliver targeted doses of radiation for cancer treatment.
-Similarly, in environmental science, the decay of isotopes such as radon is monitored to assess contamination levels and potential health risks.
+For instance, in nuclear medicine, the controlled decay of radioactive isotopes is utilized in radiotherapy to deliver targeted doses of radiation for cancer treatment (@britannica_radioactivity).
+Similarly, in environmental science, the decay of isotopes such as carbon-14 is monitored to assess contamination levels and potential health risks (@openmedscience_carbon14_decay).
 
-The stochastic approach modeled here provides critical insights into systems with low molecule counts, where randomness plays a significant role.
+The stochastic approach modeled here provides insights into systems with low molecule counts, where randomness plays a significant role.
 For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery.
-Likewise, in environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.
+In environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.
 
@@
 # Conclusion {#sec-conclusion}
-The Gillespie Algorithm represents a significant advancement in the numerical simulation of stochastic molecular reaction networks.
-Its ability to generate statistically valid trajectories of reaction systems provides critical insights into processes governed by inherent randomness, especially in systems with low molecule counts.
-This study has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.
 
-The simulation of thymine photodimerization highlighted the algorithm’s capacity to model complex reaction networks, offering a detailed view of the formation of UV-induced DNA lesions such as CPDs and 64-PPs.
+The Gillespie Algorithm represents a significant role in the numerical simulation of stochastic molecular reaction networks.
+Its ability to generate statistically trajectories of reaction systems provides insights into processes governed by inherent randomness, especially in systems with low molecule counts.
+This project has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.
+
+The [thymine photodimerization simulation](#sec-Photodimerization) illustrated the algorithm’s capacity to model complex reaction networks, offering a detailed view of the formation of UV-induced DNA lesions such as CPDs and 64-PPs.
 This has direct implications for understanding the molecular origins of UV-induced mutagenesis and skin cancer.
-Similarly, the radioactive decay simulation demonstrated the algorithm’s precision in modeling exponential decay processes and capturing stochastic fluctuations, a crucial feature for understanding radioactive isotopes in fields such as nuclear medicine and environmental science.
+Similarly, the [radioactive decay simulation](#sec-Decay) demonstrated the algorithm’s precision in modeling exponential decay processes and capturing stochastic fluctuations, a crucial feature for understanding radioactive isotopes in fields such as nuclear medicine and environmental science.
 
@@
 \section{Appendix}
-## R Code for [4.1 Photodimerization Simulation](#sec-Photodimerization)
+## R Code for [Photodimerization Simulation](#sec-Photodimerization) (section 4.1)
 
-```{r a1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
-# Load required libraries
-if (!require("reshape2")) install.packages("reshape2")
-if (!require("ggplot2")) install.packages("ggplot2")
+### Simulation
+
+```{r application-1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
+# Work environment set-up
 library(reshape2)
@@
 }
+```
+
+### Visulization for Figure \ref{fig:dimerization}
+
+```{r figure-1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
 # Run the simulation
@@
 
-## R code for [4.2 Radioactive Decay Simulation](#sec-Decay)
-```{r a2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
+## R code for [Radioactive Decay Simulation](#sec-Decay) (section 4.2)
+
+### Simulation
+
+```{r application-2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
 # Define parameters
@@
 state <- c(R = 10000)  # Initial radioactive substance
-max_time <- 1000  # Extended maximum simulation time
+max_time <- 1000
 
@@
 simulation <- simulate_gillespie(state, max_time)
+```
 
+### Visulization for Figure \ref{fig:decay-linear}
+
+```{r figure-2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
 # Load ggplot2 for visualization
@@
   theme_minimal()
+```
 
+### Visulization for Figure \ref{fig:decay-log}
+
+```{r figure-3, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
 # Logarithmic-scale plot

### 2024-11-28 paper/Gillespie-Algorithm.Rmd
Final version

--- 2024-11-28 paper/paper.Rmd
@@
 Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory).
-We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006).
+We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, and the stochastic decay of radioactive isotopes.
 These simulations will be performed using the R programming language (@citeR).
@@
 ```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
-# Load required libraries
-if (!require("reshape2")) install.packages("reshape2")
-if (!require("ggplot2")) install.packages("ggplot2")
+# Environment set-up
 library(reshape2)

### 2024-12-02 paper/Gillespie-Algorithm.Rmd
final push!
grammar edits

@@
 Omidali A.
-Jazifor for his invaluable guidance throughout this project.
+Jazi for his invaluable guidance throughout this project.
 The source code for simulations can be found on our [GitHub repository](https://github.com/masinan123/Gillespie-algorithm.git)."
 date: "`r Sys.Date()`"
-abstract: "The Gillespie Algorithm provides a framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes.
-Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics.
+abstract: "The Gillespie Algorithm provides a framework for simulating stochastic molecular reaction networks by capturing the inherent randomness of chemical and biochemical processes.
+Unlike deterministic methods, the algorithm accurately models systems with low molecular counts where fluctuations significantly influence reaction dynamics.
 This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes.
-Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes.
+These simulations highlight the algorithm’s capability to model stochastic trajectories and offer insights into reaction mechanisms and the role of randomness in molecular processes.
 The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
@@
 This algorithm was developed  to answer the question, “If a fixed volume V contains a spatially uniform mixture of N chemical species which can inter-react through M specified chemical reaction channels, then given the numbers of molecules of each species present at some initial time, what will these molecular population levels be at any later time?” (@Gillespie1977).
-Given the trajectory of this molecular reaction is a stochastic process, this algorithm can simulate various molecular equation systems, such as radioactive decay models or decaying-dimerization reaction sets (@GillespieSSA).
+Given that the trajectory of this molecular reaction is a stochastic process, this algorithm can simulate various molecular equation systems, such as radioactive decay models or decaying-dimerization reaction sets (@GillespieSSA).
 
@@
 
-In the following sections, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Section 2](#sec-Topic-Review).
-Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory).
-We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, and the stochastic decay of radioactive isotopes.
-These simulations will be performed using the R programming language (@citeR).
-Finally, our [Section 5](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.
+The following [Section 2](#sec-Topic-Review) first provides an overview of the Gillespie Algorithm and its historical development.
+Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, are detailed in [Section 3](#sec-Theory).
+In [Section 4](#sec-application), we demonstrate the application of the Gillespie Algorithm to two molecular systems: the photodimerization of thymine, and the stochastic decay of radioactive isotopes.
+These simulations are performed using the R programming language (@citeR).
+Finally, [Section 5](#sec-conclusion) summarizes our findings, discusses the algorithm's broader implications, and explores potential future applications.
 
@@
 Firstly $P(S,t)$ itself is a probabilistic description of the system.
-Next, in most of the cases, the reaction/process is stochastic.
-Therefore the master equation describes the ensemble average of all possible systems, here `ensemble` is the collection of all possible states with the same initial condition and macroscopic parameters.
+Next, in most cases, the reaction/process is stochastic.
+Therefore the master equation describes the ensemble average of all possible systems.
+In this context, `ensemble` is the collection of all possible states with the same initial condition and macroscopic parameters.
 Following some certain initial distribution, we are not guaranteed to reach the same distribution at time $t$.
 
-In the article *Cellular growth and division in the Gillespie algorithm* (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species.
+In the article *Cellular growth and division in the Gillespie algorithm* (@Lu2004), the authors discuss the system of well-stirred mixture of $N$ chemical species.
 Since the system is well-stirred, the distribution of particles is always uniform in space.
@@
 
-With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$ here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.
+With the above set-up, the master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$ here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.
 
@@
 For a system with $n$ particles, $|S_Y|$ is of order $O(n^N)$.
-Therefore it is almost impossible to solve the equation analytically, especially when the number of possible elementary reactions $M$ is large.
+It is therefore almost impossible to solve the equation analytically, especially when the number of possible elementary reactions $M$ is large.
 
@@
 Given that a reaction occurs, the probability that it is the $\mu'$-th reaction is $P(\mu=\mu')=\frac{a_{\mu'}}{A}$.
-Therefore the the original Gillespie recipe follows the following steps:
+Therefore, the the original Gillespie recipe follows the following steps:
 
@@
 
-Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T_1 \)), Thymine 2 (\( T_2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reactions channels: \( c_1 \) and \( c_2 \).
+Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T_1 \)) and Thymine 2 (\( T_2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reaction channels: \( c_1 \) and \( c_2 \).
 \[
@@
 
-We will simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per \(10^6\) normal bases) and 64-PP (~2 per \(10^6\) normal bases) found by exposing skin cells to a UVB dose of 0.2 \(J/cm^2\) (@Mouret2006).
+We simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per \(10^6\) normal bases) and 64-PP (~2 per \(10^6\) normal bases) found by exposing skin cells to a UVB dose of 0.2 \(J/cm^2\) (@Mouret2006).
 As the amount of thymine bases in DNA varies from person to person, we will simulate thymine photodimerization of 1000 pairs of thymine.
@@
 As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, which aligns with expectations based on specific UV radiation conditions and previously reported quantum yields (@MartinezFernandez2022; @Mouret2006).
-The number of reactions per unit time levels off as pyrimidine pairs decrease, which is consistent with the propensity functions, as the propensity of 64-PP formation is much significantly than the propensity of CPD formation (@Mouret2006).
-This plateau suggests that, as the number of available reactants decreases, the rate of reaction slows down.
+The number of reactions per unit time levels off as pyrimidine pairs decrease, which is consistent with the propensity functions, as the propensity of 64-PP formation is markedly higher than the propensity of CPD formation (@Mouret2006).
+This plateau suggests that as the number of available reactants decreases, the rate of reaction slows down.
 Due to the stochastic nature of molecular reactions, this simulation demonstrates how the Gillespie algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each individual reaction event.
@@
 The Gillespie algorithm offers profound implications for understanding molecular reactions, particularly in systems where stochastic effects are prominent, such as the dimerization.
-This process, highlighted in Figure \ref{fig:dimerization}, is widely studied, as it can lead to mutations potentially resulting in skin cancer, when not repaired.
-The Gillespie  algorithm allows us to realistically model these reactions, as it demonstrates the probabilistic nature of DNA damage, capturing fluctuations in molecular interactions that deterministic models cannot, providing a deeper understanding of molecular events, insights into mutation rates, and the impact of environmental factors like UV radiation on genetic integrity.
+This process, highlighted in Figure \ref{fig:dimerization}, is widely studied as it can lead to mutations potentially resulting in skin cancer when not repaired.
+The Gillespie algorithm allows us to realistically model these reactions, as it demonstrates the probabilistic nature of DNA damage by capturing fluctuations in molecular interactions that deterministic models cannot, and provides a deeper understanding of molecular events, insights into mutation rates, and the impact of environmental factors like UV radiation on genetic integrity.
 The algorithm's ability to model these stochastic processes is crucial in areas such as genetics, where random molecular events can significantly influence disease outcomes.
@@
 Using the Gillespie algorithm to model molecular reactions has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology.
-By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes, such as in disease development, medicine, and environmental impact studies.
+By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems as it enables a better prediction of outcomes, such as in disease development, medicine, and environmental impact studies.
 
@@
 
-The stochastic approach modeled here provides insights into systems with low molecule counts, where randomness plays a significant role.
+The stochastic approach modeled in this project provides insights into systems with low molecule counts, where randomness plays a significant role.
 For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery.
@@
 The Gillespie Algorithm represents a significant role in the numerical simulation of stochastic molecular reaction networks.
-Its ability to generate statistically trajectories of reaction systems provides insights into processes governed by inherent randomness, especially in systems with low molecule counts.
+Its ability to generate statistical trajectories of reaction systems provides insights into processes governed by inherent randomness, especially in systems with low molecule counts.
 This project has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.
@@
 By providing a framework for understanding and simulating stochastic reaction systems, the Gillespie Algorithm continues to play a pivotal role in advancing our knowledge of molecular biology, chemical kinetics, and physical chemistry.
-Future developments in computational efficiency and hybrid methods may further expand its applicability, enabling the simulation of even more complex systems with greater precision.
+Future developments in computational efficiency and hybrid methods may further expand its applicability and enable the simulation of even more complex systems with greater precision.
 
@@
 
-### Visulization for Figure \ref{fig:decay-linear}
+### Visulization for Figure \ref{fig:decay-linear} and Figure \ref{fig:decay-log}
 
@@
 
-### Visulization for Figure \ref{fig:decay-log}
-
 ```{r figure-3, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
